taxadb: A High-Performance Local Taxonomic Database Interface

### 2020-01-21 paper/manuscript.Rmd
Add table captions

printtable <- function(df){
  df %>%
    kableExtra::kable("latex", booktabs=T) %>%
    kableExtra::kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)

@@

Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 1), or simply using `all` to install all available providers:

@@

These issues are best illustrated using a minimal example.  We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species:

@@

```{r echo=FALSE, cache = FALSE}
status_data %>% printtable()
```

```{r echo = FALSE, cache = FALSE }
trait_data %>% printtable()
```

@@

```{r echo = FALSE, cache = FALSE}
joined %>% tidyr::replace_na(list(category = "-")) %>% printtable() # "NA" looks like a category, "-" is clearer visually
```

@@

```{r cache = FALSE, echo = FALSE}
## Just for pretty-printing
joined %>%  
  tidyr::replace_na(list(category = "-", elton_name = "-", redlist_name = "-")) %>%
  select(elton_name, redlist_name, mass, category, id) %>%
  printtable()
```

@@

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@Franz2009, @franz2018]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

```{r}
## example of sci name matching to more than one id.
filter_name("Zosterops rendovae") %>% 
  select(1:5)

@@

tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
                                 "National Center for Biological Information's Taxonomy database (NCBI)",
                                 "Catalogue of Life (COL)",
                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
                                 "FishBase (Froese and Pauly 2019)",
                                 "Open Tree Taxonomy (OTT)",
                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")

>>>>

printtable <- function(df, ...){
  df %>%
    kableExtra::kable("latex", booktabs=T, ...) %>%
    kableExtra::kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)

@@

Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 6), or simply using `all` to install all available providers:

@@

These issues are best illustrated using a minimal example.  We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species [\@ref(tab:iucn_table)]

@@

```{r iucn_table, echo=FALSE, cache = FALSE}
status_data %>% printtable(caption = "IUCN status for the selected 10 species.")
```

```{r trait_table, echo = FALSE, cache = FALSE }
trait_data %>% printtable(caption = "Trait data for the selected 10 species.")
```

@@

```{r echo = FALSE, cache = FALSE}
joined %>% tidyr::replace_na(list(category = "-")) %>% 
  printtable(caption = "IUCN and trait data joined on scientific name.") # "NA" looks like a category, "-" is clearer visually
```

@@

```{r cache = FALSE, echo = FALSE}
## Just for pretty-printing
joined %>%  
  tidyr::replace_na(list(category = "-", elton_name = "-", redlist_name = "-")) %>%
  select(elton_name, redlist_name, mass, category, id) %>%
  printtable(caption = "IUCN and trait data joined on taxonomic ID.")
```

@@

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009, @Franz2018]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

```{r}
## example of sci name matching to more than one id.
filter_name("Zosterops rendovae") %>% 
  select(1:5) %>%
  printtable(caption = "Accepted name matches for Zosterops rendovae.")

@@

tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
                                 "National Center for Biological Information's Taxonomy database (NCBI)",
                                 "Catalogue of Life (COL)",
                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
                                 "FishBase (Froese and Pauly 2019)",
                                 "Open Tree Taxonomy (OTT)",
                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                    total_identifiers = map(abbreviation,
                                            ~taxa_tbl(.x) %>%
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
                    description = desc
) %>%
  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE, 
               caption = "Descriptions of providers supported by taxaDB.") %>%
  kable_styling(latex_options="scale_down") %>%
  kableExtra::column_spec(1, width = "5cm") %>%
  kableExtra::column_spec(4, width = "10cm")
